---
title: "STA305 Final Project"
author: "Ziyi Zhang"
date: "18/12/2023"
output: 
  
  pdf_document: default
  html_document:
    df_print: paged
---

# Sleep health and Lifestyle

## Background


# Dataset Overview
The Sleep Health and Lifestyle Dataset comprises 374 rows and 13 columns, covering a wide range of variables related to sleep and daily habits. It includes details such as gender, age, occupation, sleep duration, quality of sleep, physical activity level, stress levels, BMI category, blood pressure, heart rate, daily steps, and the presence or absence of sleep disorders.

## Key Features of the Dataset
Comprehensive Sleep Metrics: Explore sleep duration, quality, and factors influencing sleep patterns.
Lifestyle Factors: Analyze physical activity levels, stress levels, and BMI categories.
Cardiovascular Health: Examine blood pressure and heart rate measurements.
Sleep Disorder Analysis: Identify the occurrence of sleep disorders such as Insomnia and Sleep Apnea.

## Dataset Columns
Person ID: An identifier for each individual.
Gender: The gender of the person (Male/Female).
Age: The age of the person in years.
Occupation: The occupation or profession of the person.
Sleep Duration (hours): The number of hours the person sleeps per day.
Quality of Sleep (scale: 1-10): A subjective rating of the quality of sleep, ranging from 1 to 10.
Physical Activity Level (minutes/day): The number of minutes the person engages in physical activity daily.
Stress Level (scale: 1-10): A subjective rating of the stress level experienced by the person, ranging from 1 to 10.
BMI Category: The BMI category of the person (e.g., Underweight, Normal, Overweight).
Blood Pressure (systolic/diastolic): The blood pressure measurement of the person, indicated as systolic pressure over diastolic pressure.
Heart Rate (bpm): The resting heart rate of the person in beats per minute.
Daily Steps: The number of steps the person takes per day.
Sleep Disorder: The presence or absence of a sleep disorder in the person (None, Insomnia, Sleep Apnea).

## Details about Sleep Disorder Column
- None: The individual does not exhibit any specific sleep disorder.
- Insomnia: The individual experiences difficulty falling asleep or staying asleep, leading to inadequate or poor-quality sleep.
- Sleep Apnea: The individual suffers from pauses in breathing during sleep, resulting in disrupted sleep patterns and potential health risks.

# Exploratory Data Analysis

List of packages:

library(tidyverse)
library(knitr)
library(kableExtra)
library(reticulate)
library(janitor)
library(latex2exp)
library(gridExtra)
library(broom)
library(patchwork)
library(crosstable)
library(agridat)
library(FrF2)
library(pwr)
library(emmeans)
library(DiagrammeR)
library(abind)
library(magic)
library(BsMD)
library(scidesignR)

```{r}
library(tidyverse)
library(latex2exp)
library(miscset)
```
### Load the data
```{r}
data <- read.csv("/Users/user/Downloads/Sleep_health_and_lifestyle_dataset.csv", header=T)

glimpse(data)
```
### Summarize the data
```{r}
summary(data)
```
It summarizes each variable in the data set.
For each of the numeric variables we can see the following information: 
  - Min: The minimum value
  - 1st Qu: The value of the first quartile (25th percentile)
  - Median: The median value
  - Mean: The mean value
  - 3rd Qu: The value of the third quartile (75th percentile)
  - Max: The maximum value
For the categorical variables in the data set (Gender, Occupation, BMI.Category, Blood.Pressure, Sleep.Disorder) we see a frequency count of each value

### Clean and sort Data
```{r}
# change Person.ID from int to chr
data$Person.ID <- as.character(data$Person.ID)

# separate Blood.Pressure into new columns "Systolic", and "Diastolic", and set Sleep Disorder as 0 if no sleep disorders, 1 otherwise
data1 <- data %>%
  separate(Blood.Pressure, c("Systolic", "Diastolic"), "/") %>%
  mutate(Sleep.Disorder = ifelse(Sleep.Disorder == "None", 0, 1))
glimpse(data1)
```
```{r} 
# change Systolic and Diastolic variables from chr to int, and Sleep.disorder into int
data1$Systolic <- as.integer(data1$Systolic)
data1$Diastolic <- as.integer(data1$Diastolic)
data1$Sleep.Disorder <- as.integer(data1$Sleep.Disorder)
```

### Numerical variables
```{r}
# Scatter plots of Numerical Variables
data1 %>%
  keep(is.numeric) %>%
  gather() %>%
  ggplot(aes(value)) +
    facet_wrap(~ key, scale = "free") +
    geom_histogram(bins=5, color="black", fill="grey")
```

### Categorical Variables
```{r, fig.height=10, fig.width=22}
# bar graph display for all the categorical variables
ggplotGrid(ncol = 2,
    lapply(c("Gender", "Occupation", "BMI.Category", "Sleep.Disorder"),
       function(col) {
         ggplot(data1, aes_string(col)) +
           geom_bar(color="black", fill="blue")
       }))
```
Analysis: 
- Gender (binary-class): Even distribution for female and male
- Occupation (multi-class): Some occupations do not have enough data plotted on the graph to draw sufficient conclusions. May consider merging those occupations into one group
- BMI.Category (multi-class): A somewhat balanced distribution for "Normal" and "Overweight", not enough data under "Obese" and "Normal Weight", considering adding it to the "Overweight", and "Normal" category
- Sleep.Disorder (multi-class): Somewhat even distributions for those with no sleep disorders and those with sleep disorders. For the categories "Sleep Apnea" and "Insomnia" that are under sleep disorder, it seems evenly distributed.

```{r, fig.height=10, fig.width=22}
# fix Occupation by combining the occupations with low data into either the category "others", or combine with a similar category of same field
data1$Occupation[data1$Occupation == 'Manager'] <- 'Others'
data1$Occupation[data1$Occupation == 'Sales Representative'] <- 'Salesperson'
data1$Occupation[data1$Occupation == 'Scientist'] <- 'Others'
data1$Occupation[data1$Occupation == 'Software Engineer'] <- 'Engineer'

# fix BMI.Category by adding "Obese" to "Overweight"
data1$BMI.Category[data1$BMI.Category == 'Obese'] <- 'Overweight'

# combine the two different labels for people in normal BMI category
data1$BMI.Category[data1$BMI.Category == 'Normal Weight'] <- 'Normal'

# show new bar graph display
ggplotGrid(ncol = 2,
    lapply(c("Gender", "Occupation", "BMI.Category", "Sleep.Disorder"),
       function(col) {
         ggplot(data1, aes_string(col)) +
           geom_bar(color="black", fill="blue")
       }))
```

```{r}
data2 <- data1 %>%
  mutate(Gender = ifelse(Gender == "Male", 0, 1)) %>%
  mutate(BMI.Category = ifelse(BMI.Category == "Normal", 0, 1)) 
glimpse(data2)
```

### Mean summary table grouped by Sleep Disorder effects
```{r}
# Define the variables used in the propensity score model
years1 <- mean(data2$Age[data2$Sleep.Disorder == 1])
years0 <- mean(data2$Age[data2$Sleep.Disorder == 0])
male1<- 100 * mean(data2$Gender[data2$Sleep.Disorder == 1] == 0)
male0 <- 100 * mean(data2$Gender[data2$Sleep.Disorder == 0] == 0)
sleeptime1 <- mean(data2$Sleep.Duration[data2$Sleep.Disorder == 1])
sleeptime0 <- mean(data2$Sleep.Duration[data2$Sleep.Disorder == 0])
activity1 <- mean(data2$Physical.Activity.Level[data2$Sleep.Disorder == 1])
activity0 <- mean(data2$Physical.Activity.Level[data2$Sleep.Disorder == 0])
stress1 <- mean(data2$Stress.Level[data2$Sleep.Disorder == 1])
stress0 <- mean(data2$Stress.Level[data2$Sleep.Disorder == 0])
normal1 <- 100 * mean(data2$BMI.Category[data2$Sleep.Disorder == 1] == 0)
normal0 <- 100 * mean(data2$BMI.Category[data2$Sleep.Disorder == 0] == 0)
rate1 <- mean(data2$Heart.Rate[data2$Sleep.Disorder == 1])
rate0 <- mean(data2$Heart.Rate[data2$Sleep.Disorder == 0])
steps1 <- mean(data2$Daily.Steps[data2$Sleep.Disorder == 1])
steps0 <- mean(data2$Daily.Steps[data2$Sleep.Disorder == 0])

df<- data.frame( T1=c(years1, male1, sleeptime1, activity1, stress1, normal1, rate1, steps1), T0 = c(years0, male0, sleeptime0, activity0, stress0, normal0, rate0, steps0))

row.n <- c("years", "gender", "sleep.duration", "physical.activity", "stress.level", "BMI.category", "heart.rate", "daily.steps")

row.names(df) <- row.n

round(df, 2)
```


```{r}
 library(Matching)
```

```{r}
mb <- MatchBalance(Sleep.Disorder ~ as.factor(Gender) + 
                     Age + Sleep.Duration + 
                     Physical.Activity.Level +
                     Stress.Level + as.factor(BMI.Category) + 
                     Heart.Rate + Daily.Steps, data = data2,nboots = 10)
```
## Propensity score matching 
```{r}
prop.model <- glm(Sleep.Disorder ~ as.factor(Gender) + 
                     Age + Sleep.Duration + Physical.Activity.Level +
                     Stress.Level + as.factor(BMI.Category) + 
                     Heart.Rate + Daily.Steps, family = binomial(), 
                    data = data2)

X <- prop.model$fitted
Y <- data2$Quality.of.Sleep
Tr <- data2$Sleep.Disorder

#head(cbind( Tr,prop.model$fitted), 20)
rr <- Match(Y = Y,Tr = Tr,X = X,M = 1)
attributes(rr)
```
```{r}
summary(rr)
```
## covariate balance
```{r}
MatchBalance(Sleep.Disorder ~ as.factor(Gender) + 
                     Age + Sleep.Duration + Physical.Activity.Level +
                     Stress.Level + as.factor(BMI.Category) + 
                     Heart.Rate + Daily.Steps, data = data2,
                     match.out = rr,nboots = 10)
```

```{r}
#Unadjusted t-test
t.test(data2$Quality.of.Sleep[as.factor(data2$Sleep.Disorder)==0],
       data2$Quality.of.Sleep[as.factor(data2$Sleep.Disorder)==1], var.equal = T)
```


## Logistic regression of Sleep Disorder on covariates 
```{r}
prop.model <- glm(Sleep.Disorder ~ as.factor(Gender) + 
                     Age + Sleep.Duration + Physical.Activity.Level +
                     Stress.Level + as.factor(BMI.Category) + 
                     Heart.Rate + Daily.Steps, family = binomial(), 
                     data = data2)

data2$pSleep.Disorderobs <- predict(prop.model, type = "response")
strat <- quantile(data2$pSleep.Disorderobs,c(.2,.4,.6,.8))

# Create strata defined by the propensity score

data2$strat1 <- data2$pSleep.Disorderobs<=strat[1]
data2$strat2 <- (data2$pSleep.Disorderobs > strat[1]) & (data2$pSleep.Disorderobs <= strat[2])
data2$strat3 <- (data2$pSleep.Disorderobs > strat[2]) & (data2$pSleep.Disorderobs <= strat[3]) 
data2$strat4 <- (data2$pSleep.Disorderobs > strat[3]) & (data2$pSleep.Disorderobs <= strat[4])
data2$strat5 <- data2$pSleep.Disorderobs > strat[4] 

data2$stratvar <- numeric(length(data2$Sleep.Disorder))

for (i in 1:length(data2$Sleep.Disorder)) 
  {
if (data2$strat1[i]==TRUE) {data2$stratvar[i] <- 1}
else 
  if (data2$strat2[i]==TRUE) {data2$stratvar[i] <- 2}
else 
  if (data2$strat3[i]==TRUE) {data2$stratvar[i] <- 3}
else 
  if (data2$strat4[i]==TRUE) {data2$stratvar[i] <- 4}
else data2$stratvar[i] <- 5
}

write.csv(data2,"data2.csv")

#The treatment effect within each strata is obtained in the R code below.


data2 <- read.csv(file = 'data2.csv')
propmodel1 <- glm(Quality.of.Sleep[strat1]~Sleep.Disorder[strat1],data = data2)

summary(propmodel1)
```
```{r}
propmodel2 <- glm(Quality.of.Sleep[strat2]~Sleep.Disorder[strat2], 
                  data = data2)
summary(propmodel2)
```
```{r}
propmodel3 <- glm(Quality.of.Sleep[strat3]~Sleep.Disorder[strat3], 
                  data = data2)
summary(propmodel3)
```
```{r}
propmodel4 <- glm(Quality.of.Sleep[strat4]~Sleep.Disorder[strat4], 
                  data = data2)
summary(propmodel4)
```
```{r}
propmodel5 <- glm(Quality.of.Sleep[strat5]~Sleep.Disorder[strat5], 
                  data = data2)

summary(propmodel5)
```
```{r}
data2 <- read.csv(file = 'data2.csv')
stratmodel <- glm(Quality.of.Sleep~Sleep.Disorder+as.factor(stratvar),data = data2)
summary(stratmodel)
```

```{r}
# 95% confidence interval for treatment effect based on subclassification
confint(stratmodel)[2,]
```

## Multivariate adjustment using the propensity score
```{r}
prop.model.adj <- glm(Quality.of.Sleep ~ Sleep.Disorder+ pSleep.Disorderobs, data = data2)
summary(prop.model.adj)
```
```{r}
confint(prop.model.adj)
```

## Multivariate adjustment using the propensity score 
## Addition of stratification
```{r}
prop.model.adjA <- glm(Quality.of.Sleep ~ Sleep.Disorder + as.factor(stratvar) +pSleep.Disorderobs , data = data2)
summary(prop.model.adjA)
```
```{r}
confint(prop.model.adjA)
```
```{r}
#prop.model.adjA <- glm(Quality.of.Sleep ~ Sleep.Disorder + as.factor(stratvar) +pSleep.Disorderobs , data = data2)
table(as.factor(data2$stratvar), data2$Sleep.Disorder )

```

